Appendix: Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-02-21 Appendix_plain_AOS.tex

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA $copies$ $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Let $\alpha_0$ be the intercept representing the mean hake DNA concentration and $\epsilon(x,y,d)$ be a random effect that allows for the effect of spatial location and water depth. We write these as additive on the log-scale,

@@

This model describes perhaps the simplest possible random effect model and notably does not leverage any information about the spatial arrangement of samples (i.e. the random effect does not include a spatial correlation component).  This is intentional.  The focus of this model is on the other species DNA concentration (not Pacific hake) and therefore we allow the estimated hake concentrations to follow the qPCR observations as much as possible.

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu L$ aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see Shelton et al. 2022 Supp S1 for additional description).

Finally, we add a random effect for the individual bottles sampled at each year-station-depth combination ($\delta_i(x,y,d)$). This effect  describes the deviation of individual sampled bottles from the location mean, $\delta_i(x,y,d) \sim N(0,\tau)$, with $\tau$ indicating a depth-specific standard deviation among bottles. Because we have only two replicate bottles at each location, we impose a sum to zero constraint on the bottles collected from a single location (i.e. for a given location, $\sum_i \delta_i(x,y,d) = 0$). Then, the log-concentration of hake DNA in a sample analyzed by qPCR, $E_{Hi}$ is,

@@

The 2 is present in the first line of the equation because we use 2 $\mu L$ of sample in each qPCR reaction and so the expected DNA copies in reaction is $2e^{E_{ij}}$. We allow $\sigma_C(E_{ij})$ to vary as a log-linear function of DNA concentration to account for the fact that there is decreased variability in qPCR measures of $C_t$ at higher DNA concentrations: $\sigma_C({E_{ij}}) = exp(\gamma_{0}+\gamma_{1}E_{ij})$ (see also Fig. XX).

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu L^{-1}$, each of 2$\mu L$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{jp}$ be the known copy number for sample $j$ in PCR plate $p$, then,

\begin{align}
  Z_{jp} &\sim Bernoulli \left(1-exp(-2 K_{jp} \phi) \right)\\
  Y_{jp} &\sim Normal \left(\beta_{0j}+\beta_{1j}\log{K_{jp}},\sigma_Y(K_{jp})\right) \quad  if \: Z_{jp} =1
\end{align}

Note that there are different intercept $\beta_{0p}$ and slope $\beta_{1p}$ parameters for each PCR plate to allow for among-plate variation in amplification. We model each calibration paratemeter ($\beta_{0p},\beta_{1p})$ hierarchically using a normal distribution, with among plate mean and variance (i.e. $\beta_{0p} \sim N(\mu_{\beta_{0}},\sigma_{\beta_{0}})$).

@@

In addition to the qPCR observations, we used the MiFish primer and high throughput sequencing to detect and enumerate the sequences of other fish species in the same samples that were examined for hake DNA. Again, we are interested in determining the DNA concentration of each focal species in each sample. Unfortunately, metabarcoding data is compositional data (see *Glor et al. *, *Shelton et al. 2023*), and so while it can inform the relative abundance of species within a sample, alone it cannot provide estimates of DNA concentration. Fortunately, the qPCR data from Pacific hake provides information about DNA concentration for one species and the metabarcoding data provides information about the relative abundance for tens of species, including hake. Here, we follow and extend the metabarcoding model developed by Shelton et al. (2023) and show how it links with the qPCR model above.

As above, we let $D_s(x,y,d)$ be the log-scale eDNA concentration but now for species $s$. We let $E_{si}(x,y,d)$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; *Aitchison XXXX*).  We define a reference species in each sample, $R$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see Shelton et al. 2023).  We can then calculate $\nu_{si}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see *Mock Communities* below),

\begin{align} 
  \nu_{si} &= \left( E_{si}(x,y,d) - E_{Ri}(x,y,d)\right) + N_{PCR} \left(\alpha_s - \alpha_R \right) 
\end{align}

To maintain identifiability, we assign one $\alpha_s = 0$. We can then convert the $\nu$s to the predicted proportions of the metabarcoding reads $\mu_{si}$ using the softmax transformation and to the observed reads via a Dirichlet-Multinomial likelihood,

@@

As with the qPCR analyses, in the absence of additional information estimates of $\alpha$ and $E$ cannot be statistically identified.  We use the mock community data, which has known input DNA concentrations and  same basic model as for the metabarcoding observations but for the moe

@@

@@

The above statistical model provides predictions of species-specific DNA concentrations (D) at each depth sampled (d = 0, 50, 150, 300 or 500 m). We used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA , as:
$$\ln D_{xyz} = \tau + \eta(s) + \varepsilon_{d}(s)$$
where x,y,z index longitude, latitude, and depth, respectively, $\tau$ is scalar intercept, $\eta(s)$ is a spatial smooth shared among depths, and $\varepsilon_d(s)$ is a depth-specific spatial smooth. Both $\eta(s)$ and $\varepsilon_d(s)$ are a zero-centered Gaussian random field estimated using the stochastic partial differntial equation (SPDE) approximation (see \cite{anderson2022}). Each species was fitted individually while we created and selected the best mesh for approximating spatial Gaussian random fields as part of the model.

\subsection*{Model convergance}
OLE + GLED

@@

\clearpage
\begin{table}\centering
\caption{Initial proportional abundances (measured by amplifying the 12S rRNA gene using MarVer1 primers \cite{valsecchi2020} and ddPCR QX200 Droplet Digital PCR system (Bio-Rad, Inc.)) of the selected species (n=12) across different mock communities (n=8). NA represent no DNA template included.}
    \begin{tabular}{lcccccccc}
        \toprule
        {Species} & \multicolumn{2}{c}{Mock1} & \multicolumn{2}{c}{Mock2} & \multicolumn{2}{c}{Mock3} & \multicolumn{2}{c}{Mock4} \\
        & Even & Skew & Even & Skew & Even & Skew & Even & Skew \\
        \midrule
        \textit{Clupea pallasii}           & 0.437 & 0.325  & 0.042  & 0.009     & -         & -          & 0.025      & 0.038 \\
        \textit{Engraulis mordax}          & 0.122 & 0.034  & 0.15 & 0.15       & 0.146      & 0.16       & 0.308      & 0.478 \\
        \textit{Leuroglossus stilbius}     & 0.108 & 0.29   & - & -         & -         & -         & -         & - \\    
        \textit{Merluccius productus}      & 0.079 & 0.144  & 0.12 & 0.06       & 0.117      & 0.075      & 0.247      & 0.287 \\
        \textit{Microstomus pacificus}     & -    & -     & 0.057  & 0.215     & -         & -         & -         & - \\    
        \textit{Sardinops sagax}           & 0.028 & 0.033  & 0.14 & 0.035      & 0.137      & 0.025      & 0.289      & 0.131 \\
        \textit{Scomber japonicus}         & -    & -     & 0.122  & 0.053      & 0.119      & 0.027     & -         & - \\    
        \textit{Sebastes entomelas}        & -    & -     & - & -          & 0.068      & 0.218     & -         & - \\    
        \textit{Stenobrachius leucopsarus} & -    & -     & - & -          & 0.356      & 0.487     & -         & - \\    
        \textit{Tactostoma macropus}       & -    & -     & 0.13 & 0.324     & -         & -         & -         & - \\    
        \textit{Tarletonbeania crenularis} & -    & -     & 0.083  & 0.124     & -         & -         & -         & - \\    
        \textit{Thaleichthys pacificus}    & 0.123 & 0.113  & 0.099  & 0.012     & -         & -          & 0.013      & 0.021 \\
        \textit{Trachurus symmetricus}     & 0.103 & 0.06   & 0.057  & 0.018      & 0.056      & 0.008      & 0.118      & 0.04 \\
        \bottomrule
    \end{tabular}
\end{table}
\clearpage
\bibliography{Bib.bib}
\end{document}

>>>>

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA $copies$ $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Hereafter we refer to a combination of spatial location and depth as a ``station". Let $\psi_0$ be the intercept representing the mean hake DNA concentration and $\epsilon(x,y,d)$ be a random effect that allows for the effect of spatial location and water depth. We write these as additive on the log-scale,

@@

This model describes perhaps the simplest possible random effect model and notably does not leverage any information about the spatial arrangement of samples (i.e. the random effect does not include a spatial correlation component).  This is intentional.  The focus of this model is on the other species DNA concentration (not Pacific hake) and therefore we allow the estimated hake concentrations at a to follow the qPCR observations as much as possible.

Unfortunately, qPCR does not directly measure DNA concentration. Instead qPCR measures the PCR cycle at which a 2$\mu L$ aliquot of a extacted DNA was detected to fluoresce and compares that observation against the flourescence pattern of samples of known DNA concentration ("the standard curve"). This provides an estimate of DNA concentration for each unknown sample. Each water sample is analyzed using at least 3 independent qPCR replicates to account for laboratory and machine variability. In addition to qPCR variability among replicates, we also have to account for modifications that affect each water sample and occurred during water sampling and processing. First, we have three offsets that modify the true DNA concentration to affect what we observed in the qPCR. For water sample $i$: 1) $V_i$ is the proportion of 2.5 L filtered from Niskin $i$ (occasionally some seawater was spilled or not filtered; $V_i =1$ for the vast majority of samples); 2) $I_i$ is the known dilution used to on sample $i$ to eliminate PCR inhibition. PCR inhibition is most commonly observed in surface samples and is vanishingly rare for samples collected below 100m; 3) $\mathbf{I}\zeta$ is an estimated offset for an ethanol wash error ($\zeta$ is the estimated effect of the wash error and $\mathbf{I}$ is an indicator variable where $\mathbf{I}=1$ for affected samples and $\mathbf{I}=0$ otherwise; see Shelton et al. 2022 Supp S1 for additional description). Here we fix $\zeta$ = XXX based on previous analyses (Shelton et al. 2022).

Finally, we add a random effect for the individual bottles sampled at each year-station-depth combination ($\delta_i(x,y,d)$). This effect  describes the deviation of individual sampled bottles from the location mean, $\delta_i(x,y,d) \sim N(0,\sigma_D)$, with $\sigma_D$ indicating a depth-specific standard deviation among bottles. Because we have only two replicate bottles at each location, we impose a sum to zero constraint on the bottles collected from a single location (i.e. for a given location, $\sum_i \delta_i(x,y,d) = 0$). Then, the log-concentration of hake DNA in a sample analyzed by qPCR, $E_{Hi}$ is,

@@

The 2 is present in the first line of the equation because we use 2 $\mu L$ of sample in each qPCR reaction and so the expected DNA copies in reaction is $2e^{E_{ij}}$. We allow $\sigma_C(E_{ij})$ to vary as a log-linear function of DNA concentration to account for the fact that there is decreased variability in qPCR measures of $C_t$ at higher DNA concentrations: $\sigma_C({E_{ij}}) = exp(\gamma_{0}+\gamma_{1}E_{ij})$.

By itself, the above model is unidentifiable because field samples do not provide information about the parameters that define relationship between the number of DNA copies and PCR cycle ($\beta_{0j}$, $\beta_{1j}$, and $\phi$). As is standard with qPCR analyses, we include standards of known concentration to estimate these parameters. Each qPCR plate has replicate samples with a known number of DNA copies. These standards span six orders of magnitude (1 to 100,000 copies $\mu L^{-1}$, each of 2$\mu L$) and determine the relationship between copy number and PCR cycle of detection. Let $K_{jp}$ be the known log-copy number for sample $j$ in PCR plate $p$, then,

\begin{align}
  Z_{jp} &\sim Bernoulli \left(1-exp(-2 e^{K_{jp}} \phi) \right)\\
  Y_{jp} &\sim Normal \left(\beta_{0j}+\beta_{1j}{K_{jp}},\sigma_Y(K_{jp})\right) \quad  if \: Z_{jp} =1
\end{align}

Note that there are different intercept ($\beta_{0p}$) and slope ($\beta_{1p}$) parameters for each PCR plate to allow for among-plate variation in amplification. We model each calibration parameter ($\beta_{0p},\beta_{1p})$ hierarchically using a normal distribution, with among plate mean and variance (i.e. $\beta_{0p} \sim N(\mu_{\beta_{0}},\sigma_{\beta_{0}})$).

@@

In addition to the qPCR observations, we used the MiFish primer and high throughput sequencing to detect and enumerate the sequences of other fish species in the same samples that were examined for hake DNA. Again, we are interested in determining the DNA concentration of each focal species in each sample. Unfortunately, metabarcoding data is compositional data \cite{glor2017, shelton2023}, and so while it can inform the relative abundance of species within a sample, alone it cannot provide estimates of DNA concentration. Fortunately, the qPCR data from Pacific hake provides information about DNA concentration for one species and the metabarcoding data provides information about the relative abundance for tens of species, including hake. Here, we follow and extend the metabarcoding model developed by Shelton et al. (2023) and show how it links with the qPCR model above.

As above, we let $D_s(x,y,d)$ be the log-scale eDNA concentration but now for species $s$. We let $E_{si}(x,y,d)$ be the log-DNA concentration for each species in each bottle. Importantly one of the $s$ species is Pacific hake and so connects to the qPCR model above. Metabarcoding reads are compositional and so we model the DNA concentration relative to one another using the additive log-ratio transformation (ALR; \textbf{Aitchison 1986}).  We define a reference species in each sample, $R$, with the species used as a reference varying among sample. Let $\alpha_s$ be the per PCR cycle log amplification rate for species $s$ (see \cite{shelton2023}).  We can then calculate $\nu_{si}$, the log-ratio between species $s$ and the reference species after accounting for amplification bias (see *Mock Communities* below),

\begin{align} 
  \nu_{si} &= \left( E_{si}(x,y,d) - E_{Ri}(x,y,d)\right) + N_{PCR} \left(\alpha_s - \alpha_{Ri} \right) 
\end{align}

We can then convert the $\nu$s to the predicted proportions of the metabarcoding reads $\mu_{si}$ using the softmax transformation and to the observed reads via a Multinomial likelihood,

@@

As with the qPCR analyses, in the absence of additional information estimates of $\alpha$ and $E$ cannot be statistically identified.  We use the mock community data, which has known input DNA concentrations and  same basic model as for the metabarcoding observations but instead of having unknown DNA concentrations, we have known inputs for each species $L_{sm}$ in mock community $m$. Then:

\begin{align} 
 \nu_{sm} &= \left( L_{sm} - L_{Rm}\right) + N_{PCR} \left(\alpha_s - \alpha_{Ri} \right) \\
  \pi_{sm} &= \dfrac{e^{\nu_{sm}}}{\sum_{s=1}^S e^{\nu_{sm}}} \\
  \boldsymbol{R}_{m} & \sim Multinomial\left(\boldsymbol{\pi}_{m},T_{m}\right) 
\end{align}

To maintain identifiability, we assign one $\alpha_s = 0$. (ADD. WHICH ONE IS ZERO?, WHICH IS THE REF SPECIES?). ADD PLOT of ALPHAS.

@@

Ole's questions:
Smoothed model is based on E (a single water bottle), not D.
What was used in the making the smooth?  Posterior mean?

@@

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). We used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA , as:
$$E_i(xyz) = \tau + \eta(s) + \varepsilon_{d}(s)$$
where x,y,z index longitude, latitude, and depth, respectively, $\tau$ is scalar intercept, $\eta(s)$ is a spatial smooth shared among depths, and $\varepsilon_d(s)$ is a depth-specific spatial smooth. Both $\eta(s)$ and $\varepsilon_d(s)$ are a zero-centered Gaussian random field estimated using the stochastic partial differntial equation (SPDE) approximation (see \cite{anderson2022}). Each species was fitted individually while we created and selected the best mesh for approximating spatial Gaussian random fields as part of the model.

\subsection*{Model convergance}

Gelman-Rubin statistics.  Predicted-observed, etc.

@@

\clearpage
\begin{table}\centering
\caption{Initial proportional abundances (measured by amplifying the 12S rRNA gene using MarVer1 primers \cite{valsecchi2020} and ddPCR QX200 Droplet Digital PCR system (Bio-Rad, Inc.)) of the selected species (n=12) across different mock communities (n=8). NA represent no DNA template included.}
    \begin{tabular}{lcccccccc}
        \toprule
        {Species} & \multicolumn{2}{c}{Mock1} & \multicolumn{2}{c}{Mock2} & \multicolumn{2}{c}{Mock3} & \multicolumn{2}{c}{Mock4} \\
        & Even & Skew & Even & Skew & Even & Skew & Even & Skew \\
        \midrule
        \textit{Clupea pallasii}           & 0.437 & 0.325  & 0.042  & 0.009     & -         & -          & 0.025      & 0.038 \\
        \textit{Engraulis mordax}          & 0.122 & 0.034  & 0.15 & 0.15       & 0.146      & 0.16       & 0.308      & 0.478 \\
        \textit{Leuroglossus stilbius}     & 0.108 & 0.29   & - & -         & -         & -         & -         & - \\    
        \textit{Merluccius productus}      & 0.079 & 0.144  & 0.12 & 0.06       & 0.117      & 0.075      & 0.247      & 0.287 \\
        \textit{Microstomus pacificus}     & -    & -     & 0.057  & 0.215     & -         & -         & -         & - \\    
        \textit{Sardinops sagax}           & 0.028 & 0.033  & 0.14 & 0.035      & 0.137      & 0.025      & 0.289      & 0.131 \\
        \textit{Scomber japonicus}         & -    & -     & 0.122  & 0.053      & 0.119      & 0.027     & -         & - \\    
        \textit{Sebastes entomelas}        & -    & -     & - & -          & 0.068      & 0.218     & -         & - \\    
        \textit{Stenobrachius leucopsarus} & -    & -     & - & -          & 0.356      & 0.487     & -         & - \\    
        \textit{Tactostoma macropus}       & -    & -     & 0.13 & 0.324     & -         & -         & -         & - \\    
        \textit{Tarletonbeania crenularis} & -    & -     & 0.083  & 0.124     & -         & -         & -         & - \\    
        \textit{Thaleichthys pacificus}    & 0.123 & 0.113  & 0.099  & 0.012     & -         & -          & 0.013      & 0.021 \\
        \textit{Trachurus symmetricus}     & 0.103 & 0.06   & 0.057  & 0.018      & 0.056      & 0.008      & 0.118      & 0.04 \\
        \bottomrule
    \end{tabular}
\end{table}
\clearpage

\begin{table}[h]
    \centering
    \begin{tabular}{lll}
         & \textbf{Description} & \textbf{Prior} \\
&Data & \\
\hline
$K$ & known DNA concentration from qPCR standards in copies/$\mu$L & - \\
$Z$ & qPCR amplification (yes=1; no=0) & - \\
$Y$ & qPCR cycle threshold (Ct) & - \\
$R$ & metabarcoding number of reads & - \\
$T$ & total number of reads for all S species & - \\
$N_{PCR}$ & number of metabarcoding PCR cycles & - \\
$L$ & known DNA concentration from mock commuinties in copies/$\mu$L & - \\
$V$ & volume filtered and aliquote dilution offset & -\\
$I$ & aliquote dilution factor & -\\

&&\\
&State processes&\\
\hline
$E$ & unknown DNA concentration in copies/$\mu$L & - \\
$D$ & unknown DNA concentration in copies/L & - \\
$F$ & smoothed DNA concentration in copies/L & - \\
&&\\

&Parameters&\\
\hline
$\psi$& mean hake eDNA concentration (copies/$\mu$L) & $\mathcal{N}(2,2)$\\
$\epsilon_{(x,y,d)}$& depth specific random effect & $\mathcal{N}(0,\tau)$\\
$\tau$& standard deviation of depth specific random effect & $\mathcal{N}(0,0.2)$\\
$\delta$& biological replicate random effect & $\mathcal{N}(0,\sigma_D)$\\
$\sigma_D$& standard deviation of biological replicate random effect  & $\mathcal{N}(0,3)$\\
$\phi$& intercept of the qPCR probability of detection relationship & $\mathcal{N}(3,1)$\\
$\beta0$& intercept of the linear relation between the mean Ct values&$\mathcal{N}(39,3)$\\
&and eDNA concentration (K and E) & \\
$\beta1$& slope of the linear relation between the mean Ct values and &$\mathcal{N}(-1.4,0.05)$\\
&eDNA concentration (K and E) & \\
$\gamma0$& intercept of the linear relation between the standard deviation&$\mathcal{N}(0,1)$\\
&of Ct values and eDNA concentration (K and E) & \\
$\gamma1$& intercept of the linear relation between the standard deviation &$\mathcal{N}(0,1)$\\
&of Ct values and eDNA concentration (K and E) & \\
$\alpha$& species amplification efficiency & fixed\\
$\zeta$& ethanol wash effect & fixed\\
&&\\
&Subscripts&\\
\hline
$s$& species ($H$ = Hake) & -\\
$x$& northing & -\\
$y$& easting & -\\
$d$& depth & -\\
$i$& biological replicate & -\\
$r$& technical replicate & -\\
$p$& qPCR plate & -\\
$m$& mock community sample & -\\
$j$& qPCR standard aliquote sample & -\\

    \end{tabular}
    \caption{Data, state processes, parameters, and subscripts employed in the joint Bayesian model and their prior distributions}
    \label{tab:filter_error}
\end{table}
\clearpage

\bibliography{Bib.bib}
\end{document}
